---
title: "Practical Machine Learning"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background
Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

## Data
The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

## Loding Data
The Dataset has been downloaded from the internet and has been loaded into two seperate dataframes, __“training__ and “testing”. The __“training__ data set has 19622 number of records and the __“testing__ data set has 20 records. The number of variables is 160.

```{r, result = 'hide', echo = T}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(RGtk2)
library(rattle)
library(randomForest)
library(gbm)
```

```{r}
init_org_training_data <- read.csv("pml-training.csv")
init_org_testing_data <- read.csv("pml-testing.csv")
dim(init_org_testing_data)
dim(init_org_training_data)
```
## Data Cleansing

### Part 1 - Remove variables which are having nearly zero variance

```{r}
non_zero_var <- nearZeroVar(init_org_training_data)


org_training_data <- init_org_training_data[,-non_zero_var]
org_testing_data <- init_org_testing_data[,-non_zero_var]

dim(org_training_data)
```

```{r}
dim(org_testing_data)

```

### Part 2 - Removing Variables which are having NA values. Our threshhold is 95%
```{r}
na_val_col <- sapply(org_training_data, function(x) mean(is.na(x))) > 0.95

org_training_data <- org_training_data[,na_val_col == FALSE]
org_testing_data <- org_testing_data[,na_val_col == FALSE]

dim(org_training_data)
```

```{r}
dim(org_testing_data)

```

### Part 3 - Removing variables which are non-numeric and hence will not contribute into our model. The very first 7 variables are of that kind only. Hence those needs to be removed from the datasets

```{r}
org_training_data <- org_training_data[,8:59]
org_testing_data <- org_testing_data[,8:59]

dim(org_training_data)
```

```{r}
dim(org_testing_data)
```

## Data Partitioning

As per recommendation of the course __ Practical Machine Learning__ , we will be seggregating our org_training_data into 2 different parts, one is the training set (consisiting 60% of the total data) and test set (consisting 40% of the total data)

```{r}
inTrain <- createDataPartition(org_training_data$classe, p=0.6, list=FALSE)
training <- org_training_data[inTrain,]
testing <- org_training_data[-inTrain,]

dim(training)
```



```{r}
dim(testing)
```

## Decision tree model
```{r}
DT_modfit <- train(classe ~ ., data = training, method="rpart")
```

Prediction in terms of Decision Tree Model

```{r}
DT_prediction <- predict(DT_modfit, testing)
confusionMatrix(DT_prediction, as.factor(testing$classe))
```

```{r}
rpart.plot(DT_modfit$finalModel, roundint=FALSE)
```

We can see that the prediction accuracy is 50% which is not upto the desired level.


### Random Forest Model

```{r}
RF_modfit <- train(classe ~ ., data = training, method = "rf", ntree = 100)
```

Prediction in terms of Random Forest Model

```{r}
RF_prediction <- predict(RF_modfit, testing)
RF_pred_conf <- confusionMatrix(RF_prediction, as.factor(testing$classe))
RF_pred_conf
```

```{r}
plot(RF_pred_conf$table, col = RF_pred_conf$byClass, 
     main = paste("Random Forest - Accuracy Level =",
                  round(RF_pred_conf$overall['Accuracy'], 4)))
```

From the Confusion Matrix, we can clearly see that the prediction accuracy of Random Forest model is 99% which is satisfactory.

## Gradient Boosting Model

```{r}
GBM_modfit <- train(classe ~ ., data = training, method = "gbm", verbose = FALSE)
GBM_modfit$finalModel
```

```{r}
#GBM_prediction <- predict(GBM_modfit, testing, type = "class", n.trees = 5, type = link)
GBM_prediction <- predict(GBM_modfit, testing)

GBM_pred_conf <- confusionMatrix(GBM_prediction, as.factor(testing$classe))
GBM_pred_conf
```

```{r}
plot(GBM_pred_conf$table, col = GBM_pred_conf$byClass, 
     main = paste("Gradient Boosting - Accuracy Level =",
                  round(GBM_pred_conf$overall['Accuracy'], 4)))
```

From Gradient Boost Model, the prediction accuracy is 95% which is satisfactory.

### Now we need to see how each model has predicted the validation dataset across the classifications. ** We are not considering Decision Tree model as it didn’t reach the satisfactory prediction accuracy level. SO only Random Forest and Gradient Boosting methods are being compared.

```{r}
RF_pred_conf$overall
```


```{r}
GBM_pred_conf$overall
```

## Conclusion

After checking the Overall Statistics data, the Random Forest model has definitely more accuracy than GBM. Hence we will be selecting Random Forest model for final prediction from org_testing_data .


## Final Prediction- Applying selected model on the Test Data

```{r}
Final_RF_prediction <- predict(RF_modfit, org_testing_data )
Final_RF_prediction
```

